hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-02-26 inst/manuscript/paper_pdf/paper.Rmd
add the new paper_pdf folder with the newest version of the paper and associated files

---
title: Template for Oxford University Press papers
author:
  - name: Morgan M. Sparks
    email: Morgan.Sparks@usda.gov
    orcid: 0000-0003-0787-2218
    corresponding_author: TRUE # Automatic footnote using corresponding_author
    affiliation: A
  - name: Eli Felts
    orcid: 0000-0002-5888-9266
    affiliation: B
  - name: Bryan M. Maitland
    orcid: 0000-0002-4491-5064
    affiliation: A
    footnote:
address:
  - code: A
    department: Rocky Mountain Research Station
    organization: U.S. Forest Service
    #postcode: 
    city: Boise
    state: Idaho # Mix and match elements to get proper address
    country: United States
  - code: B
    department: Idaho Fish and Wildlife Conservation Office
    organization: U.S. Fish and Wildlife Service
    city: Orofino
    state: Idaho # Mix and match elements to get proper address
    country: United States
abstract:
  - "Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids. Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. **hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. **hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. **hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.
"
keywords:
  - developmental phenology
  - effective value model
  - R
  - Shiny
#boxedlist: # Not fully compatible with all document formats
#  - First point
#  - Second point
competing_interests: |
  There are no competing interest.
author_contributions: |
  MMS and BMM designed and wrote the package. EF wrote the Shiny app. All authors wrote and edited the manuscript.
acknowledgments: |
  We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature datasets.

  The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS or USFWS. Any mention of trade names, products, or services does not imply an endorsement by the U.S. government, USFS, or USFWS. USFS and USFWS do not endorse any commercial products, services or enterprises.

bibliography: ms_refs.bib # Includes refs in OUP example template
## When using `citation_package = "default"`, a CSL file can be used
#csl: https://www.zotero.org/styles/oxford-university-press-note
output:
  rticles::oup_article:
    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    journal: "Fisheries"
    document_style: "contemporary" # Can be contemporary, modern, traditional
    papersize: "large" # Can be large, medium, small
    citation_package: "default" # Uncomment when using a CSL; default "natbib"
    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
    #number_sections: FALSE # Uncomment to not number sections; default TRUE
    #number_lines: TRUE # Use `lineno` package - Default FALSE
    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
    #onecolumn: TRUE # Uncomment for one column format; default FALSE
    extra_dependencies:
      - booktabs # to use with knitr::kable() example below

## Example of pandoc's variable
#urlcolor: orange
#linkcolor: green
#citecolor: red
header-includes:
  #- \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
  - \theoremstyle{thmstyleone} # Theorem stuff from OUP template
  - \newtheorem{theorem}{Theorem} #  meant for continuous numbers. %%\newtheorem{theorem}{Theorem}[section] # meant for sectionwise numbers. optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
  - \newtheorem{proposition}[theorem]{Proposition} # %%\newtheorem{proposition}{Proposition}" # to get separate numbers for theorem and proposition etc.
  - \theoremstyle{thmstyletwo}
  - \newtheorem{example}{Example}
  - \newtheorem{remark}{Remark}
  - \theoremstyle{thmstylethree}
  - \newtheorem{definition}{Definition}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, # By default, hide code; set to TRUE to see code
  fig.pos = 'th', # Places figures at top or here
  out.width = '100%', dpi = 300, # Figure resolution and size
  fig.env="figure"
) # Latex figure environment

options(knitr.table.format = "latex") # For kable tables to write LaTeX table directly
```

```{r, include=FALSE}
library(tidyverse)
library(hatchR)
library(ggridges)
```

@@

```{r echo=FALSE, message=FALSE}
### 
library(tidyverse); library(hatchR); library(ggdist)
```

@@

```{r, echo = FALSE, out.height = '75%', fig.align = 'center'}
knitr::include_graphics("flowchart.png")
```

@@

Fundamentally, **hatchR** assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade). Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required. We expect your data to look something like this:

```{r kabletable, echo=FALSE}
df = data.frame(date = c("2000-01-01",
                         "...",
                         "2000-07-01",
                         "...",
                         "2000-12-31"),
                temperature= c(2.51,
                        "...",
                        16.32,
                        "...",
                        3.13
                        ))

# kable can alse be used for creating tables
knitr::kable(df,
             booktabs=TRUE)
```

@@

```{r, echo = TRUE}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are 
                              # user customizable
                temp_max = 25)
```

@@

#### Fitting models for other fishes

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
### predict_phenology

@@

#### Understanding your results

@@

#### Plotting phenology

@@

```{r, message=FALSE, echo=FALSE}
cut_ints <- date(
  c(
    "2011-08-31", "2012-08-31",
    "2013-08-31", "2014-08-31",
    "2015-08-31"
  )
)

>>>>

---
title: ""
output: pdf_document
header-includes:
   - \usepackage{lineno}
   - \linenumbers
bibliography: ms_refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## hatchR: A toolset to predict when fish hatch and emerge

Morgan Sparks^1^, Eli Felts^2^, Allison Swartz^3^, Bryan Maitland^1^

^1^Rocky Mountain Research Station, U.S. Forest Service, Boise, ID, USA

^2^US Fish and Wildlife Service, Idaho Fish and Wildlife Conservation Office, Orofino, ID, USA

^3^College of Forestry, Oregon State University, Corvallis, OR, USA

keywords: developmental phenology, effective value model, R, Shiny

\newpage

# Abstract

Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids. Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. **hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. **hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. **hatchR** promises to open many exciting avenues in research and management of fishes during their early life history

\newpage

@@

```{r echo=FALSE, message=FALSE}
### 
library(tidyverse); library(hatchR); library(ggdist); library(ggridges)
```

@@

```{r, echo = FALSE, out.height = '75%', fig.align = 'center', fig.cap="hatchR workflow. Data inputs are represented by folders, processes by filled circles, functions as plain text rectangles, and decisions as italicized text."}
knitr::include_graphics("./workflow.png")
```

@@

Fundamentally, **hatchR** assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade). Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required. Data is expected to be formatted similar to Table 1.

```{r kabletable, echo=FALSE }
df = data.frame(date = c("2000-01-01",
                         "...",
                         "2000-07-01",
                         "...",
                         "2000-12-31"),
                temperature= c(2.51,
                        "...",
                        16.32,
                        "...",
                        3.13
                        ))

# kable can alse be used for creating tables
knitr::kable(df,
             booktabs=TRUE,
             caption = "Example temperature data for use in hatchR.")
```

@@

```{r, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check. Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are 
                              # user customizable
                temp_max = 25)
```

@@

### Fitting models for other fishes

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom built parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
### predict_phenology

@@

### Understanding your results

@@

### Plotting phenology

@@

```{r, message=FALSE, echo=FALSE,  out.width='85%', fig.align='center', fig.cap="Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data."}
cut_ints <- date(
  c(
    "2011-08-31", "2012-08-31",
    "2013-08-31", "2014-08-31",
    "2015-08-31"
  )
)
